The sweet spot for local polygenic adaptation: a case study motivated by stickleback

### 2018-09-12 Stickleback_Paper.tex
latest changes

Threespine stickleback fish provide one of the most striking examples of local adaptation. This hemisphere-wide metapopulation includes both marine populations and a large number of smaller freshwater populations
that have apparently repeatedly adapted to freshwater conditions
often using the same genetic basis.
In this paper, we use simulations motivated by stickleback populations
to examine what amounts of gene flow favor stable metapopulation polymorphism
with allele sharing,
and to further dissect the underlying dynamics.
We find that rapid, repeated adaptation 
using alleles maintained at low frequency by migration-selection balance
(the ``transporter hypothesis'')
occurs over a realistic range of intermediate rates of gene flow,
between slow, independent adaptation with low gene flow
and large migration load at high gene flow.
This is mediated mainly by the total downstream influx of alleles \emph{(maybe??)}.
We do not see evidence for strong effects encouraging genomic clustering of causal alleles
or towards particular dominance coefficients.
$F_{ST}$ scans for adaptive alleles are more likely to succeed with higher rates of gene flow,
The results support existing theory of local adaptation,
and provide a more concrete look at a particular, empirically motivated example.

@@

Migration occurs both locally along the coastline in the marine habitat
and between the marine habitat and the lakes,
and can be thought of as occurring at the juvenile stage.
The between-habitat migration rate is denoted $m$,
and will be called simply the ``migration rate''.
Each new individual in the marine habitat has freshwater parents with probability $m$;
to obtain the pair, a first parent is chosen proportional to fitness,
and a mate is chosen from the same lake as the first, also proportional to fitness.
The resulting offspring is given a spatial location in the marine habitat
at the location of the parent's lake.
Parents for a new marine individual who is not a migrant are chosen similarly (with probability $1-m$):
first, a single parent is chosen proportionally to fitness in the marine habitat,
and then a mate is chosen, 
also proportionally to fitness but reweighted by a Gaussian function 
of the distance separating the two, with stanadard deviation $1/2$. 
More specifically, if the first parent is marine individual $i$,
then marine individual $j$ is chosen as the mate
with probability proportional to $f(x_i) \exp(-2d_{ij}^2)$,
where $d_{ij}$ is the distance between the two locations.
Finally, each new marine offspring is given a position 
displaced from the first parent's position by a random Gaussian distance
with mean 0 and standard deviation 0.02, and reflected to stay within the population.
New offspring in the freshwater habitat are chosen in the same way,
except the probability that the parents are marine individuals is $m$;
any new freshwater offspring produced by marine individuals
are assigned to the lake nearest to the position of the first marine parent.

@@

@@

%%%%%%%%%%%%%%%%%
\section{Results}

% Overview
We varied the migration rate, $m$, across separate simulations
from $5 \times 10^{-5}$ to $5 \times 10^{-1}$ per individual,
i.e., between 0.01 and 100 migrants per year to and from each of the freshwater lakes.
This had a strong effect on many aspects of adaptation,
including how fast adaptation occurred in each lake,
how much alleles were shared between lakes,
and the population genetic signals left behind.
At the lowest migration rate, lakes adapted nearly independently,
while at the highest migration rate, the habitats are beginning to act like a single metapopulation
with substantial migration load.
We will now dissect what happens between these extremes.

\emph{merge with the above}
Interestingly, the ability for separate populations to locally adapt to their own selective pressure was relatively unaffected 
until the highest rate of migration between the marine and freshwater environments.
All rates of historical introgression aside from the lowest, helped both the initial and introduced freshwater populations 
share pre-existing freshwater adapted alleles.
The sharing of alleles resulted in rapid adaptation ($\approx 100$ generations) of the introduced population (split from the marine population) to adapt to the freshwater environment.
We also found that larger amounts of migration allow for more statistical power and less false positives in the resulting population genetics data ($F_{st}$ per SNP across the genome).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Local Adaptation: differentiation with gene flow}

% The progression of adaptation. How you can see introduction new alleles

Local adaptation occurred in all simulations:
as shown in Figure \ref{fig:phenotype_ts2},
freshwater and marine populations diverged
until the trait means were close to the optimal values in each habitat. 
%(Remaining simulations are shown in Figures \ref{fig:phenotype_ts1} and \ref{fig:phenotype_ts4}.)
%Create these once all plots are finalized
Greater migration decreased adaptedness only weakly
except for at the two highest migration rates (10 and 100 migrants per generation per lake),
where the mean trait value equilibrated at roughly half its optimal value and experienced panmixia, respectively.
(Figures \ref{fig:MeanPhenotype} and \ref{fig:phenotype_ts4}).
The establishment of new alleles in the lakes is visible in Figure \ref{fig:phenotype_ts2}
as jumps in the mean trait value;
in the two simulations these occur on a time scale of 100 (lower migration) to 1000 (higher migration)
generations,
\jgg{I'm not sure what this means? how do we know how long the jumps take?}
and move the trait by an amount of order 1.
Trait variation within each population was small compared to the difference between populations,
with interquartile ranges of around XXX.\plr{TODO}
Across all parameter values, differences at around 16 commonly polymorphic sites 
(eight the shift the trait in each direction)
were responsible for most of the adaptive differences between freshwater and marine habitats.
\plr{ALSO TODO: CHECK THIS}

@@

\begin{figure}
	\begin{center}
  		\includegraphics[width=1.0\linewidth]{semi_final_plots/Fst_semi_final.png}
  		\caption{
		Finish this when we finalize plots!
		 }
  		\label{fig:Fst}
	\end{center}
\end{figure}

\paragraph*{Speed of adaptation}
Increasing migration rate strongly decreased the time until freshwater habitats could adapt,
both in the initial and introduced sets of lakes.
As shown in Figure \ref{fig:TimeToAdaptation}, at the lowest migration rate 
it took over $20,000$ generations for the mean trait value across introduced lakes to 
get to within 0.5 of the original lakes value.
Although many lakes had adapted before this time,
the different rates of introduction of effect alleles is clearly seen in the traces
of trait values against time (e.g., Figure \ref{fig:phenotype_ts2}).
However, higher migration rates allowed the freshwater habitats to adapt
nearly as quickly as the marine habitat.

\begin{figure}
	\begin{center}
  		\includegraphics[width=1.0\linewidth]{semi_final_plots/T_adapt_semi_final.png}
  		\caption{Time to adaptation as a function of migration rate ($M$) parameter value. This is where we measure how many generation
		it takes for the introduced population's mean phenotype to come within 0.5 of the original lakes average phenotype. 
		Each point represents a simulation run at some value of $M$. 
		The yellow dashed line is the average of all points at each respective parameter value}
  		\label{fig:TimeToAdaptation}
	\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Sharing of freshwater adapted alleles}

%FIGURES THAT SHOW THIS:
%-MPAA / IND
%-Total/Avg shared

\begin{figure}
	\begin{center}
  		\includegraphics[width=\linewidth]{semi_final_plots/FAA_semi_final.png}
  		\caption{
			finish this after plots are finalized
		}
		\label{fig:MPFAI}
	\end{center}
\end{figure}

@@

\paragraph{Standing genetic variation}
\jgg{todo, make this figure for the supp}
\jgg{this will need updated in I make the plot for it}
Despite a dramatic difference in the amount of allelic sharing between lakes,
standing genetic variance (Figure \ref{fig:SGV}) 
was around 0.05 in the freshwater habitat
across all but the highest migration rates.
Concurrently,
genetic variation in the marine population steady increased with migration
to a similar value.
On the one hand, it is not surprising that lakes, as a group,
carry more genetic variation than the marine habitat,
since population subdivision allows different alleles to become common in each.
However, it is striking that at $m=.005$, the marine habitat carries as much genetic variation,
despite a lack of any substantial migration load.

@@

With 10 effect regions across the genome, we can see in \ref{fig:Fst3} that only about 1 or 2 of the region contain peaks. 
%SOFT STATEMENT - MORE (Peak) EVALUATION?
When looking closely at the peaks we see that each is usually a composition of 7-8 SNPs close in proximity.
%SOFT STATEMENT -
Because some of those mutations are selectively neutral (we can see this because the counts of FAA are not as high as the number in the clusters),
this is suggestive of hitchhiking alleles along with the SNPs being brought to high frequency by a selective sweeps.

@@

%The lowest migration rate of $10^{-5}$ also has a very low false positive rate across all $F_{st}$ threshold values.
%\subsubsection*{Low Recombination Causes Clustering}

>>>>

Threespine stickleback fish provide one of the most striking examples of local adaptation. This hemisphere-wide metapopulation includes both marine populations and a large number of smaller freshwater populations
that have apparently repeatedly adapted to freshwater conditions
often using the same genetic basis.
In this paper, we use simulations motivated by stickleback populations
to examine what amounts of gene flow favor stable metapopulation polymorphism
with allele sharing,
and to further dissect the underlying dynamics.
We find that rapid, repeated adaptation 
using alleles maintained at low frequency by migration-selection balance
(the ``transporter hypothesis'')
occurs over a realistic range of intermediate rates of gene flow,
between slow, independent adaptation with low gene flow
and large migration load at high gene flow.
This is mediated mainly by the total downstream influx of alleles \emph{(maybe??)(No)}.
We do not see evidence for strong effects encouraging genomic clustering of causal alleles
or towards particular dominance coefficients.
$F_{ST}$ scans for adaptive alleles are more likely to succeed with higher rates of gene flow,
The results support existing theory of local adaptation,
and provide a more concrete look at a particular, empirically motivated example.

@@

Migration occurs both locally along the coastline in the marine habitat
and between the marine habitat and the lakes,
and can be thought of as occurring at the juvenile stage.
The between-habitat migration rate is denoted $m$,
and will be called simply the ``migration rate''.
Each new individual in the marine habitat has freshwater parents with probability $m$;
to obtain the pair, a first parent is chosen proportional to fitness,
and a mate is chosen from the same lake as the first, also proportional to fitness.
The resulting offspring is given a spatial location in the marine habitat
at the location of the parent's lake.
Parents for a new marine individual who is not a migrant are chosen similarly (with probability $1-m$):
first, a single parent is chosen proportionally to fitness in the marine habitat,
and then a mate is chosen, 
also proportionally to fitness but reweighed by a Gaussian function 
of the distance separating the two, with standard deviation $1/2$. 
More specifically, if the first parent is marine individual $i$,
then marine individual $j$ is chosen as the mate
with probability proportional to $f(x_i) \exp(-2d_{ij}^2)$,
where $d_{ij}$ is the distance between the two locations.
Finally, each new marine offspring is given a position 
displaced from the first parent's position by a random Gaussian distance
with mean 0 and standard deviation 0.02, and reflected to stay within the population.
New offspring in the freshwater habitat are chosen in the same way,
except the probability that the parents are marine individuals is $m$;
any new freshwater offspring produced by marine individuals
are assigned to the lake nearest to the position of the first marine parent.

@@

\paragraph{Using SLiM's TreeSeq}
This case study involves analyzing large simulations which model populations under the influence of selection. 
To fully understand the origins of adaptive variants, calculate statistics, and avoid the computationally expensive task of 
simulating neutral mutations, we have used SLiM's \emph{treeSeq} feature to track the genealogical history of all samples throughout the simulation. 
This feature outputs a compact representation the genomic outcome of every meiosis relevant to tracked samples (Often thought of as the ARG).
\jgg{maybe the below should be in descriptive statistics}
To understand recomposition of the freshwater haplotype in the new lakes through selection, 
we were able to classify each adaptive variant, in each genome in the population at the generation at time of adaptation,
as deriving from one of four origins:
(1) Migrants, brought in from a migrant after the introduction of the new lakes 
(2) Pre-existing, passed down from an individual in the original generation of the new lakes, 
however, not defined to be an adaptive variant in the original lakes
(3) Captured, Pre-existing, passed down from an individual in the original generation of the new lakes, 
and defined to be adaptive in the original lakes
(4) De novo, mutated and propagated from an individual existing in the introduced lakes population, after introduction.

We also used the tree sequence to count the "fitness" of individuals in the initial generation of the new lakes. 
The "fitness" of an individual in this case, is defined to be the total number off eventual offspring, in the generation at the time of adaptation, 
which have inherited a adaptive variant from that individual. 
To calculate this, we only considered gene trees which overlapped an adaptive variant site (defined at time of adaptation).
We then counted the total number of offspring which had inherited from any adaptive variant site, 
for each individual in the initial generation.

@@

\section{Results}

We varied the migration rate, $m$, across separate simulations
from $5 \times 10^{-5}$ to $5 \times 10^{-1}$ per individual,
i.e., between 0.01 and 100 migrants per year to and from each of the freshwater lakes.
This had a strong effect on many aspects of adaptation,
including how fast adaptation occurred in each lake,
how much alleles were shared between lakes,
and the population genetic signals left behind.
At the lowest migration rate, lakes adapted nearly independently,
while at the highest migration rate, the habitats are beginning to act like a single metapopulation
with substantial migration load.
We will now dissect what happens between these extremes.
Interestingly, all rates of $m$ aside from the two highest had little affect on populations ability to locally adapt.
All rates of historical introgression aside from the lowest, helped both the initial and introduced freshwater populations 
share pre-existing freshwater adapted alleles.
The sharing of alleles resulted in rapid adaptation ($\approx 60$ generations) of the introduced population (split from the marine population) to adapt to the freshwater environment.
We also found that larger amounts of migration allow for more statistical power and less false positives in the resulting population genetics data ($F_{st}$ per SNP across the genome).

\subsection*{Local Adaptation: differentiation with gene flow}

\jgg{This needs some work}
Local adaptation occurred in all simulations:
as shown in Figure \ref{fig:phenotype_ts2},
freshwater and marine populations diverged
until the trait means were close to the optimal values in each habitat. 
The establishment of new alleles in the lakes is visible in Figure \ref{fig:phenotype_ts2}
as jumps in the mean trait value;
in the two simulations these occur on a time scale of 100 (lower migration) to 1000 (higher migration) generations,
\jgg{I'm not sure what this means? how do we know how long the jumps take?}
and move the trait by an amount of order 1.
Trait variation within each population was small compared to the difference between populations,
with interquartile ranges of around XXX.\plr{TODO}
Across all parameter values, differences at around 16 commonly polymorphic sites 
(eight the shift the trait in each direction)
were responsible for most of the adaptive differences between freshwater and marine habitats.
\plr{ALSO TODO: CHECK THIS}

@@

\begin{figure}
	\begin{center}
  		\includegraphics[width=1.0\linewidth]{semi_final_plots/Fst_semi_final.png}
  		\caption{
		$F_{st}$ as a function of genomic location between all combinations of the three subpopulations; 
		the old lakes, the new lakes, and the marine.
		$F_{st}$ was calculated based upon allele frequencies in genomic windows of 500 mb. 
		The four columns of subplot represent separate simulations with increasing $m$. 
		The three rows represent the combinations of the three separate populations.
		Ten vertical dotted pink lines in each subplot represent regions which have the potential to affect phenotype
		 }
  		\label{fig:Fst}
	\end{center}
\end{figure}

\paragraph*{Speed of adaptation}
Increasing migration rate strongly decreased the time until freshwater habitats could adapt,
both in the old and new sets of lakes.
The different rates of introduction of effect alleles is clearly seen in the traces
of trait values against time (e.g., Figure \ref{fig:phenotype_ts2}).
As shown in Figure \ref{fig:TimeToAdaptation}, at the lowest migration rate 
it took over $20,000$ generations for the mean trait value across introduced lakes to 
get to within 0.5 of the original lakes value.
At $m = 1.0$, we saw the introduced population adapt in just 32 generations. 
This rate of adaptation follows what we've observed in natural populations of Threespine stickleback.
At the highest migration rate, $m = 100$, all individuals were unable to locally adapt.

\begin{figure}
	\begin{center}
  		\includegraphics[width=1.0\linewidth]{semi_final_plots/T_adapt_semi_final.png}
  		\caption{
		Time to adaptation as a function of migration rate ($M$) parameter value. This is where we measure how many generation
		it takes for the introduced population's mean phenotype to come within 0.5 of the original lakes average phenotype. 
		Each point represents a simulation run at some value of $M$. 
		The yellow dashed line is the average of all points at each respective parameter value}
  		\label{fig:TimeToAdaptation}
	\end{center}
\end{figure}

\subsection*{Sharing of freshwater adapted alleles}

\begin{figure}
	\begin{center}
  		\includegraphics[width=\linewidth]{semi_final_plots/FAA_semi_final.png}
  		\caption{
		Distribution of freshwater adapted alleles 
		}
		\label{fig:MPFAI}
	\end{center}
\end{figure}

@@

@@

With 10 effect regions across the genome, we can see in \ref{fig:Fst3} that only about 1 or 2 of the region contain peaks. 
When looking closely at the peaks we see that each is usually a composition of 7-8 SNPs close in proximity.
Because some of those mutations are selectively neutral (we can see this because the counts of FAA are not as high as the number in the clusters),
this is suggestive of hitchhiking alleles along with the SNPs being brought to high frequency by a selective sweeps.

@@

\begin{figure}
	\begin{center}
  		\includegraphics[width=0.7\linewidth]{semi_final_plots/OriginRatio.png}
  		\caption{ 
		Ratio of Origins of adaptive alleles in all genomes at time of local adaptation for different values of $m$
		}
  		\label{fig:Origin}
	\end{center}
\end{figure}

%The lowest migration rate of $10^{-5}$ also has a very low false positive rate across all $F_{st}$ threshold values.
%\subsubsection*{Low Recombination Causes Clustering}
\subsubsection*{Origin of introduced freshwater adaptive alleles: a conundrum}
Up to this point, we have quantified rapid and parallel adaptation and found strong evidence that sharing of alleles is (more than likely?) responsible.
However in Figure \ref{fig:MPFAI} at $m$ equal to $0.1$ \emph{mig/gen/lake}, we see sharing of alleles between populations,
yet \emph{Time to adaptation} is still fairly slow ($\approx 2,000$ gens).
This is surprising when considering we see a similar quantity of allele sharing at  $m$ equal to $1.0$ \emph{mig/gen/lake}, 
but much more rapid adaptation ($\approx 60$ gens). 
Initially, this 30X speedup of adaptation might seem to be the result of subsequent migration from the marine after the point of freshwater introduction. 
This theory would assert that the continued influx of alleles is necessary for rapid local adaptation at the rate which we've observed in Nature.
However, in Figure \ref{fig:Origin}, we find that at both $m = 0.1$ and $m = 1.0$, the vast majority of adaptive alleles origins were found to propagate 
from individuals in the original generation ("Captured" OR "Pre-Existing").

The alternative explanation is that at a higher migration rate, a higher saturation of marine adapted alleles are linked to freshwater adapted alleles.
This linkage makes rebuilding the freshwater haplotype, in the new lakes, much more difficult as it much wait for recombination to break the links.
With more historical introgression between the two selective pressures, the freshwater haplotype would remain more "intact" and much more efficient
to rebuild in the lake.

%Now lets make this more concrete
